---
title: "The Anatomy of a Hit: Statistically Learning from the Best"
subtitle: "An attempt to model"
author: 
  - Luca Carnegie
thanks: "Code and data are available at: https://github.com/lcarnegie/popularity-modeling. Special thanks go to Rajan Maghera, Oliver Cao, and Hannah Yu for the support and encouragement. You're all wonderful."
date: April 18 2024
date-format: long
abstract: "We attempt to do some interesting things with models and music - stay tuned!"
format: pdf
toc: true
number-sections: true
bibliography: references.bib
nocite: |
  @*
---

```{r}
#| include: false
#| warning: false
#| message: false

library(tidyverse)
library(spotifyr)


```

\newpage

# Introduction

Throughout the history of the music industry, musicians successful or not have always pined for one thing - writing their first hit and making it big. However, for large music firms, taking on musicians that have yet to generate considerable popularity is a large risk, since they are often gambling on an unknown je ne sais quoi, and often, any given artist is more likely to not produce a hit than produce a hit; this is easily shown through the presence of a few small "superstars" that consistently dominate the music market [@rosen1981]. This poses an interesting question for artists vying to become that superstar: what does make a song a "hit"? 

One way to measure how popular a song is through the 'valence' or perceived energy of a song.[investigate the motivation here]. Measuring valence has been done at a national and cross-artistic-work level [@dodds2010], but I investigate solely song valence on an individual basis. Though difficult to exactly predict what individual song could be a hit, it is very possible to wean out insight from previously released music to statistically learn the most important elements of a great song. 

Using data from Spotify's API, I construct a linear model of valence or "energy", using the discographies of the ten most popular artists [find a source] of every particular genre in every decade since 1950. I find that [there are interesting things to be found...]. Blah Blah Blah. 

In the streaming era, competition for stardom is at an all-time high. Being cognizant of the most impactful elements could encourage artists to focus on what truly works to generate hits. This may allow them to carve their own paths outside of the controlling nature of a record contract [@burke1997]. These insights could be used to give more knowledge and freedom to artists to be able to strike out on their own and carve their own destinies in the world of music. 



# Data {#sec-data}

## Workflow {#sec-first-point}
## Source
## Measurement and Sampling 
## Variables of Interest 
### Energy

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```

### Valence

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```

### Danceability

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```

### Explicit Status

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```

### Loudness

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```


### Duration 

```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```



# Model

Create a Correl matrix to motivate choosing of variables. 

## Model set-up


### Model justification


# Results

I guess put the results of the model?

# Discussion

What does everything mean?

## First discussion point {#sec-first-point}

## Second discussion point {#sec-second-point}

## Third discussion point {#sec-third-point}

## Weaknesses and next steps {#sec-weaknesses-next}
- Standard errors be huge. That means the regression is very imprecise. One way to
resolve that would be to get a larger sample size. 
- Sample size is kind of small - but that's the point? 
Maybe constructing a fuller dataset with the entire discographies 
of these people using Spotify's assistance would allow us to get more precise
estimates of effects. 
- People could repeat this experiment on a genre-based basis or using different 
ratings of artists. Perhaps there are genre-specific relationships within other 
variables in the spotify API that I haven't covered here. 


\newpage


# References

::: {#refs}
:::

\newpage

\appendix

# Appendix {-}

# Model Testing  






```{r boilerplate-r-chunk}
#| message: false
#| echo: false
#| warning: false
#| fig-cap: CAPTION
#| fig-align: ALIGNMENT


```